---
title: "PCA batch RNAseq corrected datasets (Kwok & NC)
output: html_notebook
---

```{r}
library(DESeq2)
library(tidyverse)
library(sva)
library(ggplot2)
library(ggrepel)
library(ggfortify)
library(BiocParallel)
```

```{r}
load("CLLdata/Kwok_data/rc_merged.Rdata")
load("CLLdata/Kwok_data/metadata_merged.Rdata")
```

Run ComBat_seq on raw count data

```{r}
rc_NC_Kwok_merged <- as.matrix(rc_NC_Kwok_merged)
corrected_data = ComBat_seq(counts = rc_NC_Kwok_merged, batch = metadata_merged$batch, group = NULL)
corrected_data <- as.data.frame(corrected_data)
```

DESeq2 normalization and vst

```{r}
## Create DESeq2Dataset object
dds <- DESeqDataSetFromMatrix(countData = corrected_data, colData = metadata_merged, design = ~ CLLtype)
```

```{r}
#Pre-filter the genes which have low counts
dds <- dds[rowSums(counts(dds)) >= 10,]
```

```{r}
#Generate normalized counts
dds <- estimateSizeFactors(dds)
dds.vst<-DESeq2::varianceStabilizingTransformation(dds)
exp_merged_corrected <- SummarizedExperiment::assay(dds.vst)
exp_merged_corrected <- as.data.frame(exp_merged_corrected)
```

PCA plot

```{r}
expMerged_corrected <- exp_merged_corrected
expMerged_corrected <- t(expMerged_corrected)

pca <- prcomp(expMerged_corrected)
summary(pca)
autoplot(pca, data = expMerged_corrected)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
